Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2023-07-20 ba_resubmit/kundinger_final.tex
suggested edits to sec 3.

\usepackage{lineno}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue,filecolor=blue,backref=page]{hyperref}
\usepackage{graphicx}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}
\newtheorem{rem}{Remark}
\usepackage{booktabs}       % professional-quality tables
\usepackage{algorithm}      % algorithm environment
\usepackage{algpseudocode}
\usepackage{subcaption}

@@

\textcolor{red}{For \text{fabl}, we also use (\ref{eqn:fs_model}) and (\ref{eqn:m_dist}), along with Dirichlet prior distributions for the $\mathbf{m}$ and $\mathbf{u}$ parameters.  However, instead of the prior over $\bm{Z}$}
%Instead of the prior over $\bm{Z}$ 
from \cite{sadinle_bayesian_2017}, we follow \cite{wortman2019} and use independent priors for each component $Z_j$. However, unlike \cite{wortman2019} who proposes a flat prior for $Z_j$, 
\textcolor{teal}{we propose an alternative, called the fast beta linkage (fabl) prior.}
%the fast Beta prior as follows. 
For each $Z_j$, we have
\begin{subequations}
\begin{align}
	p(Z_j = q| \pi) &= \begin{cases} 
	\frac{1}{n_A}\pi,  & q \leq n_A;\\
	1-\pi, &  q  = n_A + j;  \\
\end{cases} \label{eqn:fast_beta_prior}  \\
\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi}) \label{eqn:fast_beta_prior2}.
\end{align}
\end{subequations}
We can interpret (\ref{eqn:fast_beta_prior}) as follows: record $B_j$ has some match in $A$ with probability $\pi$, and each record $A_i$ is equally likely to be that match. The hyperparameters $\alpha_{\pi}$ and $\beta_{\pi}$ encode prior beliefs about  the proportion of records in $B$ that have matches in $A.$

@@

\begin{subequations}
\begin{align}
	\mathcal{L}(\bm{Z}, \bm{m}, \bm{u} \mid \gamma) &= \prod_{i=1}^{n_A}  \prod_{j=1}^{n_B}\prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[  m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}, \label{eqn:likelihood}\\
	\bm{m}_f &\sim \text{Dirichlet}(\alpha_{f1}, \ldots, \alpha_{f L_f}),\quad  \forall  \; f = 1, \ldots, F, \label{eqn:m} \\
	\bm{u}_f &\sim \text{Dirichlet}(\beta_{f1}, \ldots, \beta_{f L_f}), \quad \forall \; f = 1, \ldots, F,  \label{eqn:u}\\
	p(Z_j = q| \pi)  &=
	\begin{cases} 
		\frac{1}{n_A}\pi,  & q \leq n_A; \\
		1-\pi, &  q  = n_A + j; \\
	\end{cases} \label{eqn:z}\\
	\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi})\label{eqn:pi},
\end{align}
\end{subequations}
\textcolor{teal}{where recall that $\alpha_{\pi}, \beta_{\pi}$ are fixed hyper-parameters.}

\begin{rem}
In the \cite{wortman2019} flat prior, each value $\{1, \ldots, n_A, n_A +j\}$ is a priori equally likely for $Z_j$. This amounts to a prior probability of $n_A / (n_A + 1)$ that record $B_j$ has a match in $A$. In our preliminary studies, the this highly informative prior weighting on matching can result in overly optimistic match rates. Hence, we prefer (\ref{eqn:fast_beta_prior}). We also note that the flat prior is equivalent to a special case of the fast Beta prior with $\pi$ fixed at the mean of a $\text{Beta}\left(1, 1 / n_A \right)$ random variable.
\end{rem}

\begin{rem}
Linkage with \texttt{fabl} is conducted at the record level, rather than at the record pair level, as in the Fellegi-Sunter model. That is, $\pi$ under \texttt{fabl} estimates the proportion of records in $B$ that have matches, whereas $\lambda$ in the Fellegi-Sunter model estimates the proportion of record pairs that are matches. In the bipartite case, we conjecture that some analysts will find $\pi$ in (\ref{eqn:fast_beta_prior2}) to be more a interpretable parameter than $\lambda$ in (\ref{eqn:lambda_dist}). In this setting, there are at most $n_B$ matching pairs out of $n_A n_B$ total pairs, meaning that $\lambda$ is bounded above by $\frac{1}{n_A}$ and tends towards 0 as the size of the linkage task grows. Additionally, while the Fellegi-Sunter model makes $n_A  n_B$ independent matching decisions and \texttt{BRL} makes $n_B$ dependent matching decisions, \texttt{fabl} strikes a middle ground between the two, making $n_B$  independent matching decisions. As shown in Sections \ref{sec:simulations} and \ref{sec:case-studies}, this allows \texttt{fabl} to fit a Bayesian record linkage model like \texttt{BRL} while making computational efficiency gains possible by exploiting independence. 
\end{rem}

@@

\textcolor{teal}{We present a Gibbs sampler to approximate the joint posterior distribution of $\bm{Z}, \bm{m},$ and  $\bm{u}$ given the observed comparison data from the likelihood and priors assumed in (\ref{eqn:likelihood} - \ref{eqn:pi}). While we could marginalize over $\pi$ and derive a collapsed Gibbs sampler, in the spirit of \cite{sadinle_bayesian_2017}, this leads to sequential updates, which are computationally slow. Instead, we proceed with a traditional Gibbs sampler, where the updates to the matching label (or coreference matrix) are parallel, allowing for more efficient updates. Refer to Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}

We initialize $\bm{m}$ and $\bm{u}$ from random draws from their prior distributions, and initialize $\bm{Z}$ to reflect no matches across data files; that is, $\bm{Z} = (n_A + 1, \ldots, n_A + n_B)$. \textcolor{red}{Need to initialize all random variables, such as $\pi$.} \textcolor{teal}{Denote the current draw of the sampler by $s$ and the updated draw by $s+1$.} \textcolor{teal}{For all $f=1, \ldots, F$ sample}

@@

Next, the full conditional for $\pi$ is given by
\begin{align}
\label{eqn:pi}
	\pi^{(s+1)}|\bm{Z}^{(s)},  \alpha_{\pi}, \beta_{\pi}, \pi \sim  \text{Beta}(n_{AB}(\bm{Z}^{(s)}) + \alpha_{\pi}, n_B - n_{AB}(\bm{Z}^{(s)}) + \beta_{\pi}).
\end{align}
where $n_{AB}(\bm{Z}^{(s)}) = \sum_{j=1}^{n_B} I(Z_j^{(s)} \leq n_A)$.

@@

\textcolor{teal}{
Observe that the full conditional in (\ref{eqn:z_full_conditional}) does not depend on $\bm{Z}^{(s)},$ implying that 
\begin{align}
p(\bm{Z}^{(s+1)} \mid \gamma, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)}) \propto \prod_q \left(\prod_j
p\left(Z_j^{(s+1)}  = q \mid \gamma, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)} \right)
\right).
\end{align}
}

\textcolor{teal}{This implies that we can make parallel updates to $\bm{Z}$ instead of the traditional sequential updates as in \cite{sadinle_bayesian_2017}.
In fact, to our knowledge, our approach and that of \cite{wortman2019} are the only two Bayesian approaches that propose parallel updates to the matching label (or linkage structure), contrasting the Bayesian record linkage literature \citep{tancredi2011hierarchical, sadinle_detecting_2014, steorts_bayesian_2016, marchant_distributed_2019, zanella_flexible_2016, betancourt2022prior, aleshinguendel2021multifile}. }
%See Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}

@@

\textcolor{teal}{Finally, to obtain an estimate $\hat{\bm{Z}}$ of the linkage structure, we use the loss functions and Bayes estimate from \cite{sadinle_bayesian_2017}. However, since (\ref{eqn:fast_beta_prior}) does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $B$ to one record in $A$. To obtain a Bayes estimate that fulfills the one-to-one matching (or bipartite) requirement, we minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. See Appendix \ref{bayes-estimate} for details regarding the initial Bayes estimate and the post-processing procedure.}

>>>>

\usepackage{lineno}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue,filecolor=blue,backref=page]{hyperref}
\usepackage{graphicx}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}
\newtheorem{rem}{Remark}
\usepackage{booktabs}       % professional-quality tables
\usepackage{algorithm}      % algorithm environment
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{soul}

@@

\textcolor{red}{For \text{fabl}, we also use (\ref{eqn:fs_model}) and (\ref{eqn:m_dist}), along with Dirichlet prior distributions for the $\mathbf{m}$ and $\mathbf{u}$ parameters.  However, instead of the prior over $\bm{Z}$}
%Instead of the prior over $\bm{Z}$ 
from \cite{sadinle_bayesian_2017}, we follow \cite{wortman2019} and use independent priors for each component $Z_j$. However, unlike \cite{wortman2019} who proposes a flat prior for $Z_j$, 
\textcolor{teal}{we propose an alternative, called the fast beta linkage (fabl) prior.}
%the fast Beta prior as follows. 
For each $Z_j$, we have

\begin{subequations}
\begin{align}
	p(Z_j = q| \pi) &= \begin{cases} 
	\frac{1}{n_A}\pi,  & q \leq n_A;\\
	1-\pi, &  q  = n_A + j;  \\
\end{cases} \label{eqn:fast_beta_prior}  \\
\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi}) \label{eqn:fast_beta_prior2}.
\end{align}
\end{subequations}
We can interpret (\ref{eqn:fast_beta_prior}) as follows: record $B_j$ has some match in $A$ with probability $\pi$, and each record $A_i$ is equally likely to be that match. The hyperparameters $\alpha_{\pi}$ and $\beta_{\pi}$ encode prior beliefs about  the proportion of records in $B$ that have matches in $A.$

\begin{rem}
\textcolor{red}{
In the \cite{wortman2019} flat prior, each value $\{1, \ldots, n_A, n_A +j\}$ is a priori equally likely for $Z_j$. This amounts to a prior probability of $n_A / (n_A + 1)$ that record $B_j$ has a match in $A$. In our preliminary studies, this highly informative prior weighting on matching can result in overly optimistic match rates. Hence, \textcolor{red}{we prefer (\ref{eqn:fast_beta_prior}).} We also note that the flat prior is equivalent to a special case of the fast Beta prior with $\pi$ fixed at the mean of a $\text{Beta}\left(1, 1 / n_A \right)$ random variable.
}
\end{rem}

\begin{rem}
Linkage with \texttt{fabl} is conducted at the record level, rather than at the record pair level, as in the Fellegi-Sunter model. That is, $\pi$ under \texttt{fabl} estimates the proportion of records in $B$ that have matches, whereas $\lambda$ in \textcolor{red}{the \cite{fellegi_theory_1969} model} estimates the proportion of record pairs that are matches. In the bipartite case, we conjecture that some analysts will find $\pi$ in (\ref{eqn:fast_beta_prior2}) to be more a interpretable parameter than $\lambda$ in (\ref{eqn:lambda_dist}). In this setting, there are at most $n_B$ matching pairs out of $n_A n_B$ total pairs, meaning that $\lambda$ is bounded above by \textcolor{red}{$1/n_A$} and tends towards 0 as the size of the linkage task grows. Additionally, while the Fellegi-Sunter model makes $n_A  n_B$ independent matching decisions and \texttt{BRL} makes $n_B$ dependent matching decisions, \texttt{fabl} strikes a middle ground between the two, making $n_B$  independent matching decisions. As shown in Sections \ref{sec:simulations} and \ref{sec:case-studies}, this allows \texttt{fabl} to fit a Bayesian record linkage model like \texttt{BRL} while making computational efficiency gains possible by exploiting independence. 
\end{rem}

@@

\begin{subequations}
\begin{align}
	\mathcal{L}(\bm{Z}, \bm{m}, \bm{u} \mid \gamma) &= \prod_{i=1}^{n_A}  \prod_{j=1}^{n_B}\prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[  m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}, \label{eqn:likelihood}\\
	\bm{m}_f &\sim \text{Dirichlet}(\alpha_{f1}, \ldots, \alpha_{f L_f}),\quad  \forall  \; f = 1, \ldots, F, \label{eqn:m} \\
	\bm{u}_f &\sim \text{Dirichlet}(\beta_{f1}, \ldots, \beta_{f L_f}), \quad \forall \; f = 1, \ldots, F,  \label{eqn:u}\\
	p(Z_j = q| \pi)  &=
	\begin{cases} 
		\frac{1}{n_A}\pi,  & q \leq n_A; \\
		1-\pi, &  q  = n_A + j; \\
	\end{cases} \label{eqn:z}\\
	\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi})\label{eqn:pi-prior},
\end{align}
\end{subequations}
\textcolor{teal}{where recall that \textcolor{red}{$(\alpha_{\pi}, \beta_{\pi})$} are fixed hyper-parameters.}

@@

\textcolor{teal}{We \textcolor{red}{now} present a Gibbs sampler to approximate the joint posterior distribution of $\bm{Z}, \bm{m},$ $\bm{u},$ and $\pi$ given the observed comparison data from the likelihood and priors assumed in (\ref{eqn:likelihood} - \ref{eqn:pi-prior}). 
%While we could marginalize over $\pi$ and derive a collapsed Gibbs sampler, in the spirit of \cite{sadinle_bayesian_2017}, this leads to sequential updates, which are computationally slow. Instead, we proceed with a traditional Gibbs sampler, where the updates to the matching label (or coreference matrix) are parallel, allowing for more efficient updates.
Refer to Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}
%%
We initialize $\bm{m}$ and $\bm{u}$ from random draws from their prior distributions, and initialize $\bm{Z}$ to reflect no matches across data files; that is, $\bm{Z} = (n_A + 1, \ldots, n_A + n_B)$. \textcolor{blue}{Brian: state how you initialize $\pi$.} \textcolor{teal}{Denote the current draw of the sampler by $s$ and the updated draw by $s+1$.} \textcolor{red}{To update $\mathbf{m}$ and $\mathbf{u}$, for all, $f=1, \ldots, F,$ sample}

@@

\textcolor{red}{Next, we sample $\pi$ from its full conditional, given by}
%Next, the full conditional for $\pi$ is given by
\begin{align}
\label{eqn:pi}
	\pi^{(s+1)}|\bm{Z}^{(s)},  \alpha_{\pi}, \beta_{\pi}, \pi \sim  \text{Beta}(n_{AB}(\bm{Z}^{(s)}) + \alpha_{\pi}, n_B - n_{AB}(\bm{Z}^{(s)}) + \beta_{\pi}).
\end{align}
where $n_{AB}(\bm{Z}^{(s)}) = \sum_{j=1}^{n_B} I(Z_j^{(s)} \leq n_A)$.

@@

\textcolor{teal}{
The full conditional in (\ref{eqn:z_full_conditional}) does not depend on $\bm{Z}^{(s)},$ implying that 
\begin{align}
p(\bm{Z}^{(s+1)} \mid \gamma, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)}) \propto \prod_q \left(\prod_j
p\left(Z_j^{(s+1)}  = q \mid \gamma, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)} \right)
\right).
\end{align}
}

\textcolor{teal}{Thus, we can make parallel updates to $\bm{Z}$ instead of sequential updates as in \cite{sadinle_bayesian_2017}.
In fact, to our knowledge, our approach and that of \cite{wortman2019} are the only two Bayesian approaches that propose parallel updates to the matching label (or linkage structure), \textcolor{red}{in contrast to} \textcolor{red}{other} Bayesian record linkage \textcolor{red}{methods} \citep{tancredi2011hierarchical, sadinle_detecting_2014, steorts_bayesian_2016, marchant_distributed_2019, zanella_flexible_2016, betancourt2022prior, aleshinguendel2021multifile}. }
%See Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}

@@

\textcolor{teal}{Finally, to obtain an estimate $\hat{\bm{Z}}$ of the linkage structure, we use the loss functions and Bayes estimate from \cite{sadinle_bayesian_2017}. Since (\ref{eqn:fast_beta_prior}) does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $B$ to one record in $A$. To obtain a Bayes estimate that fulfills the \textcolor{red}{bipartite} requirement, we minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. See Appendix \ref{bayes-estimate} for details regarding the initial Bayes estimate and \textcolor{red}{this} post-processing procedure.}
